### 2015-06-27 latex/paper-PLOS_ONE-revised.tex
Final version as published.

\begin{flushleft}
{\Large\textbf{Maximum allowed solvent accessibilites of residues in proteins}}
\\
\mbox{}\\
Matthew Z. Tien$^1$,
Austin G. Meyer$^{2,3}$,
Dariya K. Sydykova$^{2}$,
Stephanie J. Spielman$^{2}$,
Claus O. Wilke$^{2,\ast}$
\\
\mbox{}\\
\textbf{1} Dept.\ of Biochemistry \& Molecular Biology, The University of Chicago, Chicago, IL 60637, USA
\\
\textbf{2} Section of Integrative Biology, Institute for Cellular and Molecular Biology, and Center for Computational Biology and Bioinformatics, The University of Texas at Austin, Austin, TX 78731, USA
\\
\textbf{3} School of Medicine, Texas Tech University Health Sciences Center, Lubbock, TX, 79430, USA
\\
$\ast$ Email: wilke@austin.utexas.edu
\end{flushleft}

@@

\subsection*{Modeling Tripeptides Yields Significantly Higher Maximum ASA Values}
To derive maximum ASA values for each amino acid X, we computationally constructed Gly-X-Gly tripeptides and systematically rotated them through all biophysically allowed  conformations (see Methods and Supporting Text for details.) When constructing the tripeptides, we set bond lengths and angles (excluding $\omega$, $\phi$, $\psi$, and $\chi$ angles) for each amino acid equal to the average values observed for that amino acid in our reference set of 3197 PDB structures. We set $\omega=180^\circ$. We then rotated the $\phi$ and $\psi$ around the X residue in  discrete 1$^\circ$ steps, exhaustively enumerating all conformations. Additionally, we iterated through all rotamer angles $\chi$ that were sterically possible with each $(\phi, \psi)$ combination. For those amino acids with more than 10 possible distinct rotamer conformations, as determined by the Dunbruck database \cite{WangDunbrack2003}, we evaluated ten randomly chosen rotamer conformations. We recorded the maximum ASA observed for each $(\phi, \psi)$ backbone-angle combination.

Next, we compared the resulting theoretical maximum ASA values to the empirically observed maximum ASA values. We binned both the theoretical and the empirical values into discrete $5^\circ\times5^\circ$ bins of $(\phi, \psi)$ and recorded the maximum ASA in each bin. To eliminate nonexistent or rare conformations, we defined four Ramachandran regions for each amino acid: CORE, containing at least 80\% of the empirical observations; ALLOWED, containing at least 97\% of the empirical observations; GENEROUS, extending the core region by $20^\circ$ in all directions; and ALL, containing all non-empty bins. The definitions of the CORE, ALLOWED, and GENEROUS regions are consistent with the definitions used in Ref.~\cite{Morrisetal1992}. For each region, we displayed the maximum ASA value in each bin in side-by-side Ramachandran plots (Figures~\ref{fig:heatrama} and \ref{fig:Ala_Rama_HSV_ALL_edited}-\ref{fig:Ala_Rama_HSV_CORE_edited}) and generally found good congruence between the theoretical and the empirical values for all amino acids. Regions that had the highest maximum ASA in the theoretical data set also had the highest maximum ASA in the empirical data set. The highest ASA values were generally observed in the $\alpha$-helix region of the Ramachandran plot (Figure~\ref{fig:heatrama}). Based on these results, we propose new maximum ASA values (Tables~\ref{tab:results} and \ref{tab:max_SA_complete}) and maximally exposed geometries for each amino acid (Table~\ref{tab:max_exposed_RAMA}).

@@

We next derived two scales based on mean RSA, calculated using either our theoretical or our empirical ASA normalization values (Table~\ref{tab:hydro_scales}). Both of our mean RSA scales correlated well with the Rose scale ($r=0.96$ and $r=0.97$, respectively, with $P<10^{-10}$ in both cases) but were not identical to it. The biggest difference arose for histidine, which is ranked as the 8th-most hydrophobic amino acid according to the Rose scale but as the 10th- or 13th-most hydrophobic amino acid, respectively, according to our scales. Our scales correlated more strongly than the Rose scale with all experimental scales except the Mac Callum scale, which did not correlate significantly with either our or the Rose scale (Table~\ref{tab:Corr}). For the majority of experimental scales, the percent variance explained increased by approximately 10 percentage points using our normalization over the Rose normalization. We can conclude from these results that mean RSA is a useful measure of amino acid hydrophobicity and that correct ASA normalization is required to assign appropriate hydrophobicity scores to all amino acids.

One concern with using mean RSA as a measure of hydrophobicity is that the RSA distribution of individual amino acids tends to be highly skewed (see Figure~\ref{fig:Alanine_RSA_distribution} for an example). Hence, mean RSA may not accurately reflect the most common RSA values. It might be preferable to use instead the fraction of times an amino acid occurs in a buried conformation in empirical protein structures. This approach was originally suggested by Chothia \emph{et al.} in 1976 and executed with the limited data available at the time~\cite{Chothia1976}.

@@

Our method of obtaining ASA normalization values was similar to the methods employed by Rose \emph{et al.}~\cite{Rose1985} and by Miller \emph{et al.}~\cite{Miller1987}. Rose \emph{et al.}~\cite{Rose1985} calculated their ASA normalization values by computing the ASA of residue X in Gly-X-Gly tripeptides whose conformations were chosen based on the average dihedral angles from available empirical data at the time. Miller \emph{et al.}~\cite{Miller1987}, on the other hand, calculated their ASA normalization values by computing the ASA of an extended trimer structure with $\phi = -120^\circ, \psi= 140^\circ$ and with side-chain conformations that were frequently observed in the empirical data. The key distinction between these previous approaches and ours lies in our exhaustive sampling of tripeptide conformations. By modeling all biophysically feasible discrete combinations of $\phi$ and $\psi$ angles and varying rotamers, we identified the ideal conformations which yield maximum allowed ASA. To pursue our modeling strategy, we developed a program that allowed us to easily construct peptide chains from scratch in arbitrary conformations (see Supporting Text for details).

@@

We next filtered by allowed Ramachandran angles. For each amino acid, we binned all observed $\phi, \psi$ combinations into $5^\circ\times5^\circ$ squares, and assigned each square to one or more of the following regions: The CORE region was defined to contain at least 80\% of the observed Ramachandran angles. The ALLOWED region was defined to contain at least 97\% of the observed Ramachandran angles. For both the CORE and the ALLOWED regions, we identified, for each amino acid, the number of observations per $5^\circ\times5^\circ$ bin required for that bin to be part of the respective region. Table~\ref{tab:bin_cutoffs} lists these bin cutoffs. The GENEROUS region was defined to extend the ALLOWED region by $20^\circ$ in all directions, regardless of whether the particular Ramachandran angles have been observed. Finally, the ALL region was defined to contain all observed Ramachandran angles. The definitions of the CORE, ALLOWED, and GENEROUS regions are consistent with current biochemical convention \cite{Morrisetal1992,Laskowskietal1993}. For all four regions, we identified the maximum ASA observed.

@@

To find the theoretical maximum solvent accessibility (ASA) for each amino acid X, we computationally constructed Gly-X-Gly tripeptides. Each tripeptide was modeled by specifying coordinates of each constituent atom, using bond lengths and angles from our empirically mined protein structures. Briefly, we first constructed peptides in a defined conformation by placing each atom at the correct position in 3D space. We then adjusted $\phi$, $\psi$, and $\chi$ angles to obtain the desired conformation. This method is described in more detail in Supporting Text, and the computer code to carry out tripeptide construction has been published as a stand-alone library \cite{Tienetal2013}.

@@

@@

\begin{table}[H]
\caption{\label{tab:Corr}Absolute value of correlation coefficients $r$ between empirically derived and experimentally derived hydrophobicity scales. The largest significant correlation in each row is highlighted in bold.}
\begin{center}
\footnotesize
\begin{tabular}{lccccc}
 & \multicolumn{5}{c}{Empirical scale} \\ \cline{2-6}\\[-2ex]
Experimental scale & Mean RSA (Rose)$^\text{a}$ & Mean RSA (theor)$^\text{b}$ & Mean RSA (emp)$^\text{c}$ & 100\% buried$^\text{d}$ & 95\% buried$^\text{e}$\\
\hline
Wolfenden \emph{et al.}$^\text{f}$ & 0.614 & 0.681 & 0.681 & \textbf{0.827} & 0.774 \\
Kyte \& Doolittle$^\text{g}$ & 0.841 & 0.879 & 0.881 & \textbf{0.953} & 0.948 \\
Radzicka \& Wolfenden$^\text{h}$ & 0.852 & 0.855 & 0.851 & 0.844 & \textbf{0.888} \\
Moon \& Fleming$^\text{i}$ & 0.704 & 0.748 & 0.752 & 0.678 & \bf{0.764} \\
Fauchere \& Pliska$^\text{l}$ & 0.904 & 0.906 & \textbf{0.910} & 0.734 & 0.878 \\
Wimley \emph{et al.}$^\text{m}$ & 0.463$^\dagger$ & 0.464 & \textbf{0.473} & 0.323$^\dagger$ & 0.417$^\dagger$ \\
MacCallum \emph{et al.}$^\text{k}$ & 0.27$^\dagger$ & 0.265$^\dagger$ & 0.285$^\dagger$ & 0.116$^\dagger$ & 0.227$^\dagger$ \\
\hline
\end{tabular}
\end{center}
{\footnotesize
\begin{raggedright}
$^\text{a}$Mean RSA of residues in protein structures, as calculated by Rose \emph{et al.}~\cite{Rose1985}.\\
$^\text{b}$Mean RSA of residues in protein structures, as given in column 2 of Table~\ref{tab:hydro_scales}.\\ 
$^\text{c}$Mean RSA of residues in protein structures, as given in column 3 of Table~\ref{tab:hydro_scales}.\\
$^\text{d}$Fraction of 100\% buried residues, as given in column 4 of Table~\ref{tab:hydro_scales}.\\
$^\text{e}$Fraction of 95\% buried residues, as given in column 5 of Table~\ref{tab:hydro_scales}.\\
$^\text{f}$Transfer energy from vapor to water~\cite{Wolfenden1981}.\\
$^\text{g}$Hybrid scale based on transfer energy from vapor to water and on the percentages of 95\% and 100\% buried residues in protein structures~\cite{Kyte1981}.\\
$^\text{h}$Transfer energy from cyclohexane to water~\cite{Radzicka1988}.\\
$^\text{i}$$\Delta\Delta G$ between the folded and unfolded state of a mutated membrane-inserted protein, outer membrane phospholipase~A~\cite{Moon2011}. \\
$^\text{k}$Transfer energy calculated from molecular-dynamic simulations of side-chain analogs within a bilayer~\cite{MacCallum2007}.\\
$^\text{l}$Transfer energy between octanol and water~\cite{Fauchere1983}.\\
$^\text{m}$Transfer energy of pentapeptides between octanol and water~\cite{Wimley1996}.\\
$^\dagger$Correlation not statistically significant; all other correlations are significant at $\alpha=0.05$.\\
\end{raggedright}
}
\end{table}

>>>>

\begin{flushleft}
{\Large\textbf{Maximum allowed solvent accessibilites of residues in proteins}}
\\
\mbox{}\\
Matthew Z. Tien$^1$,
Austin G. Meyer$^{2,3}$,
Dariya K. Sydykova$^{3,4,5}$,
Stephanie J. Spielman$^{3,4,5}$,
Claus O. Wilke$^{3,4,5\ast}$
\\
\mbox{}\\
\textbf{1} Deptartment of Biochemistry \& Molecular Biology, The University of Chicago, Chicago, Illinois, United States of America
\\
\textbf{2} School of Medicine, Texas Tech University Health Sciences Center, Lubbock, Texas, United States of America
\\
\textbf{3} Institute for Cellular and Molecular Biology, The University of Texas at Austin, Austin, Texas, United States of America
\\
\textbf{4} Center for Computational Biology and Bioinformatics, The University of Texas at Austin, Austin, Texas, United States of America
\\
\textbf{5} Department of Integrative Biology, The University of Texas at Austin, Austin, Texas, United States of America
\\
$\ast$ Email: wilke@austin.utexas.edu
\end{flushleft}

@@

\subsection*{Modeling Tripeptides Yields Significantly Higher Maximum ASA Values}
To derive maximum ASA values for each amino acid X, we computationally constructed Gly-X-Gly tripeptides and systematically rotated them through all biophysically allowed  conformations (see Methods and Supporting Text in File S1 for details.) When constructing the tripeptides, we set bond lengths and angles (excluding $\omega$, $\phi$, $\psi$, and $\chi$ angles) for each amino acid equal to the average values observed for that amino acid in our reference set of 3197 PDB structures. We set $\omega=180^\circ$. We then rotated the $\phi$ and $\psi$ around the X residue in  discrete 1$^\circ$ steps, exhaustively enumerating all conformations. Additionally, we iterated through all rotamer angles $\chi$ that were sterically possible with each $(\phi, \psi)$ combination. For those amino acids with more than 10 possible distinct rotamer conformations, as determined by the Dunbruck database \cite{WangDunbrack2003}, we evaluated ten randomly chosen rotamer conformations. We recorded the maximum ASA observed for each $(\phi, \psi)$ backbone-angle combination.

Next, we compared the resulting theoretical maximum ASA values to the empirically observed maximum ASA values. We binned both the theoretical and the empirical values into discrete $5^\circ\times5^\circ$ bins of $(\phi, \psi)$ and recorded the maximum ASA in each bin. To eliminate nonexistent or rare conformations, we defined four Ramachandran regions for each amino acid: CORE, containing at least 80\% of the empirical observations; ALLOWED, containing at least 97\% of the empirical observations; GENEROUS, extending the core region by $20^\circ$ in all directions; and ALL, containing all non-empty bins. The definitions of the CORE, ALLOWED, and GENEROUS regions are consistent with the definitions used in Ref.~\cite{Morrisetal1992}. For each region, we displayed the maximum ASA value in each bin in side-by-side Ramachandran plots (Figure~\ref{fig:heatrama} and Figures \ref{fig:Ala_Rama_HSV_ALL_edited}-\ref{fig:Ala_Rama_HSV_CORE_edited} in File~S1) and generally found good congruence between the theoretical and the empirical values for all amino acids. Regions that had the highest maximum ASA in the theoretical data set also had the highest maximum ASA in the empirical data set. The highest ASA values were generally observed in the $\alpha$-helix region of the Ramachandran plot (Figure~\ref{fig:heatrama}). Based on these results, we propose new maximum ASA values (Table~\ref{tab:results} and Table~\ref{tab:max_SA_complete} in File~S1) and maximally exposed geometries for each amino acid (Table~\ref{tab:max_exposed_RAMA} in File~S1).

@@

We next derived two scales based on mean RSA, calculated using either our theoretical or our empirical ASA normalization values (Table~\ref{tab:hydro_scales} in File~S1). Both of our mean RSA scales correlated well with the Rose scale ($r=0.96$ and $r=0.97$, respectively, with $P<10^{-10}$ in both cases) but were not identical to it. The biggest difference arose for histidine, which is ranked as the 8th-most hydrophobic amino acid according to the Rose scale but as the 10th- or 13th-most hydrophobic amino acid, respectively, according to our scales. Our scales correlated more strongly than the Rose scale with all experimental scales except the Mac Callum scale, which did not correlate significantly with either our or the Rose scale (Table~\ref{tab:Corr}). For the majority of experimental scales, the percent variance explained increased by approximately 10 percentage points using our normalization over the Rose normalization. We can conclude from these results that mean RSA is a useful measure of amino acid hydrophobicity and that correct ASA normalization is required to assign appropriate hydrophobicity scores to all amino acids.

One concern with using mean RSA as a measure of hydrophobicity is that the RSA distribution of individual amino acids tends to be highly skewed (see Figure~\ref{fig:Alanine_RSA_distribution} in File~S1 for an example). Hence, mean RSA may not accurately reflect the most common RSA values. It might be preferable to use instead the fraction of times an amino acid occurs in a buried conformation in empirical protein structures. This approach was originally suggested by Chothia \emph{et al.} in 1976 and executed with the limited data available at the time~\cite{Chothia1976}.

@@

Our method of obtaining ASA normalization values was similar to the methods employed by Rose \emph{et al.}~\cite{Rose1985} and by Miller \emph{et al.}~\cite{Miller1987}. Rose \emph{et al.}~\cite{Rose1985} calculated their ASA normalization values by computing the ASA of residue X in Gly-X-Gly tripeptides whose conformations were chosen based on the average dihedral angles from available empirical data at the time. Miller \emph{et al.}~\cite{Miller1987}, on the other hand, calculated their ASA normalization values by computing the ASA of an extended trimer structure with $\phi = -120^\circ, \psi= 140^\circ$ and with side-chain conformations that were frequently observed in the empirical data. The key distinction between these previous approaches and ours lies in our exhaustive sampling of tripeptide conformations. By modeling all biophysically feasible discrete combinations of $\phi$ and $\psi$ angles and varying rotamers, we identified the ideal conformations which yield maximum allowed ASA. To pursue our modeling strategy, we developed a program that allowed us to easily construct peptide chains from scratch in arbitrary conformations (see Supporting Text in File S1 for details).

@@

We next filtered by allowed Ramachandran angles. For each amino acid, we binned all observed $\phi, \psi$ combinations into $5^\circ\times5^\circ$ squares, and assigned each square to one or more of the following regions: The CORE region was defined to contain at least 80\% of the observed Ramachandran angles. The ALLOWED region was defined to contain at least 97\% of the observed Ramachandran angles. For both the CORE and the ALLOWED regions, we identified, for each amino acid, the number of observations per $5^\circ\times5^\circ$ bin required for that bin to be part of the respective region. Table~\ref{tab:bin_cutoffs} in File~S1 lists these bin cutoffs. The GENEROUS region was defined to extend the ALLOWED region by $20^\circ$ in all directions, regardless of whether the particular Ramachandran angles have been observed. Finally, the ALL region was defined to contain all observed Ramachandran angles. The definitions of the CORE, ALLOWED, and GENEROUS regions are consistent with current biochemical convention \cite{Morrisetal1992,Laskowskietal1993}. For all four regions, we identified the maximum ASA observed.

@@

To find the theoretical maximum solvent accessibility (ASA) for each amino acid X, we computationally constructed Gly-X-Gly tripeptides. Each tripeptide was modeled by specifying coordinates of each constituent atom, using bond lengths and angles from our empirically mined protein structures. Briefly, we first constructed peptides in a defined conformation by placing each atom at the correct position in 3D space. We then adjusted $\phi$, $\psi$, and $\chi$ angles to obtain the desired conformation. This method is described in more detail in Supporting Text (File S1), and the computer code to carry out tripeptide construction has been published as a stand-alone library \cite{Tienetal2013}.

@@

\section*{Supporting Information}

File S1. Combined pdf of Figures S1--S4, Tables S1--S4, and Supporting Text.

@@

\begin{table}[H]
\caption{\label{tab:Corr}Absolute value of correlation coefficients $r$ between empirically derived and experimentally derived hydrophobicity scales. The largest significant correlation in each row is highlighted in bold.}
\begin{center}
\footnotesize
\begin{tabular}{lccccc}
 & \multicolumn{5}{c}{Empirical scale} \\ \cline{2-6}\\[-2ex]
Experimental scale & Mean RSA (Rose)$^\text{a}$ & Mean RSA (theor)$^\text{b}$ & Mean RSA (emp)$^\text{c}$ & 100\% buried$^\text{d}$ & 95\% buried$^\text{e}$\\
\hline
Wolfenden \emph{et al.}$^\text{f}$ & 0.614 & 0.681 & 0.681 & \textbf{0.827} & 0.774 \\
Kyte \& Doolittle$^\text{g}$ & 0.841 & 0.879 & 0.881 & \textbf{0.953} & 0.948 \\
Radzicka \& Wolfenden$^\text{h}$ & 0.852 & 0.855 & 0.851 & 0.844 & \textbf{0.888} \\
Moon \& Fleming$^\text{i}$ & 0.704 & 0.748 & 0.752 & 0.678 & \bf{0.764} \\
Fauchere \& Pliska$^\text{l}$ & 0.904 & 0.906 & \textbf{0.910} & 0.734 & 0.878 \\
Wimley \emph{et al.}$^\text{m}$ & 0.463$^\dagger$ & 0.464 & \textbf{0.473} & 0.323$^\dagger$ & 0.417$^\dagger$ \\
MacCallum \emph{et al.}$^\text{k}$ & 0.27$^\dagger$ & 0.265$^\dagger$ & 0.285$^\dagger$ & 0.116$^\dagger$ & 0.227$^\dagger$ \\
\hline
\end{tabular}
\end{center}
{\footnotesize
\begin{raggedright}
$^\text{a}$Mean RSA of residues in protein structures, as calculated by Rose \emph{et al.}~\cite{Rose1985}.\\
$^\text{b}$Mean RSA of residues in protein structures, as given in column 2 of Table~\ref{tab:hydro_scales} in File~S1.\\ 
$^\text{c}$Mean RSA of residues in protein structures, as given in column 3 of Table~\ref{tab:hydro_scales} in File~S1.\\
$^\text{d}$Fraction of 100\% buried residues, as given in column 4 of Table~\ref{tab:hydro_scales} in File~S1.\\
$^\text{e}$Fraction of 95\% buried residues, as given in column 5 of Table~\ref{tab:hydro_scales} in File~S1.\\
$^\text{f}$Transfer energy from vapor to water~\cite{Wolfenden1981}.\\
$^\text{g}$Hybrid scale based on transfer energy from vapor to water and on the percentages of 95\% and 100\% buried residues in protein structures~\cite{Kyte1981}.\\
$^\text{h}$Transfer energy from cyclohexane to water~\cite{Radzicka1988}.\\
$^\text{i}$$\Delta\Delta G$ between the folded and unfolded state of a mutated membrane-inserted protein, outer membrane phospholipase~A~\cite{Moon2011}. \\
$^\text{k}$Transfer energy calculated from molecular-dynamic simulations of side-chain analogs within a bilayer~\cite{MacCallum2007}.\\
$^\text{l}$Transfer energy between octanol and water~\cite{Fauchere1983}.\\
$^\text{m}$Transfer energy of pentapeptides between octanol and water~\cite{Wimley1996}.\\
$^\dagger$Correlation not statistically significant; all other correlations are significant at $\alpha=0.05$.\\
\end{raggedright}
}
\end{table}
